---
title: "Differential Protein Expression Analysis"
author: "Vishnu Ramasubramanian"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,root.dir = "C:/Users/ivanv/Desktop/UMN_Projects/Ritesh_DESeq_Analysis/TripleNullMutant_Paper_DEP/ProteinQuant/")

```

# Introduction
This document outlines the workflow for differential enrichment analysis of Mass Spectrometer (MS) proteomics data using the **DEP** package in R. Three comparisons are analyzed:

1. **MO vs. TNB**
2. **Bert vs. TN (CRISPR)**
3. **Bert vs. TN1 (CRISPR)** & **Bert vs. TN2 (CRISPR)**

For the second comparison , TN lines WPT673.7.8.8 & WPT673.7.12.5 are considered the same TN line with 4 replicates for comparison with Bert. Whereas for comparison 3, TN lines WPT673.7.8.8 & WPT673.7.12.5 are considered as distinct with 2 replicates each for comparison with Bert.

# Objective

The objective of this analysis is to compare the protein expression profile of‘conventionally-bred triple-mutant lines_vs_recurrent parents’ and ‘crispr triple-mutants_vs_bert’. This analysis will help us assess crispr tech and conventionally generated mutant lines in terms of changes in the proteome landscape induced by these two different approaches to generating mutants.

# Data Processing for DEP

## Load Required Packages & Data

```{r load-packages}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("DEP", quietly = TRUE)) BiocManager::install("DEP")
library(DEP)
library(SummarizedExperiment)
library(ggplot2)
library(ggrepel)

```

### **Analysis: MO vs. TNB**

### Load Data: Import MS2 protein level data for the samples generated using Spectronaut analysis from '.csv' files 

```{r load-data-mo-tnb}

setwd("C:/Users/ivanv/Desktop/UMN_Projects/Ritesh_DESeq_Analysis/TripleNullMutant_Paper_DEP/ProteinQuant/")

PrtnQuant <- read.csv("Supplemental table X1 Breeding lines TN .csv", header=TRUE)
dim(PrtnQuant)
```

### **Analysis: Bert vs. TN (CRISPR)**

### Load Data:  Import MS2 protein level data for the samples generated using Spectronaut analysis from '.csv' files 

```{r load-data-bert-tn}

setwd("C:/Users/ivanv/Desktop/UMN_Projects/Ritesh_DESeq_Analysis/TripleNullMutant_Paper_DEP/ProteinQuant/")

PrtnQuant2 <- read.csv("Supplemental table X2 CRISPR TN .csv", header=TRUE)
dim(PrtnQuant2)


```


## Data Preprocessing for MO vs TNB

```{r preprocess-mo-tnb}
PrtnQuant_Filt <- PrtnQuant[-c(1, which(duplicated(PrtnQuant[, 1]))), 1:5]
colnames(PrtnQuant_Filt)[1] <- "ID"

# Create sample annotation
treatment_conditions <- c("Control", "Control", "Mutant", "Mutant")
replicates <- c(1, 2, 1, 2)
sample_anno <- data.frame(label = colnames(PrtnQuant_Filt)[2:5], condition = treatment_conditions, replicate = replicates)

PrtnQuant_Filt <- make_unique(PrtnQuant_Filt, names="ID", ids="ID")
PrtnQuant_Filt[, 2:5] <- lapply(PrtnQuant_Filt[, 2:5], as.numeric)
```

### Create Summarized Experiment Object

```{r create-se-mo-tnb}
data_se <- make_se(proteins_unique = PrtnQuant_Filt, columns = c(2:5), expdesign = sample_anno)
```

### Quality checks for presence of proteins

```{r qualityChecks }

plot_frequency(data_se)
plot_numbers(data_se)




```

### Normalization using 'vsn' method 

Normalize data and visualize mean-SD relationship for nearly horizontal mean SD line. In addition, compare distributions of pre-normalized and svn-normalized data

```{r normalize-mo-tnb}

data_se_norm <- normalize_vsn(data_se)
meanSdPlot(data_se_norm)

plot_normalization(data_se, data_se_norm)

```


### Filtering


```{r filter_mo_vs_tnb}

data_se_norm_filt <- filter_missval(data_se_norm, thr = 0.5)




```

### Missing value imputation 

```{r plot-filtered_mo_vs_tnb}

plot_missval(data_se_norm)
plot_detect(data_se_norm)


plot_missval(data_se_norm_filt)
plot_detect(data_se_norm_filt)

```

### imputation 
```{r impute_mo_vs_tnb}

#For missing at random ('MAR') data a Knn or MLE
# For missing not at random ('MNAR'),MinProb (QRILC)

# data_se_norm_imp0 <- impute(data_se_norm_filt, fun = "MinProb", q = 0.01)

data_se_norm_imp1 <- impute(data_se_norm_filt, fun = "knn", rowmax=0.9)

plot_imputation(data_se_norm_imp1)

```

## Data Preprocessing for Bert vs Crispr TN


```{r preprocess-bert-tn}
PrtnQuant2_Filt <- PrtnQuant2[-c(1, which(duplicated(PrtnQuant2[, 1]))), 1:9]

# Ensure no duplicates and filter relevant columns

dim(PrtnQuant2_Filt)  # Check dimensions after filtering

colnames(PrtnQuant2_Filt)[1] <- "ID"
colnames(PrtnQuant2_Filt)[2:5] <- gsub("Biological.", "", colnames(PrtnQuant2_Filt)[2:5])
colnames(PrtnQuant2_Filt)[2:5] <- gsub("\\..Control.*.", "", colnames(PrtnQuant2_Filt)[2:5])




PrtnQuant2_Filt <- make_unique(PrtnQuant2_Filt,names="ID",ids="ID")
for(i in 2:9){
  PrtnQuant2_Filt[,i] <- as.numeric(PrtnQuant2_Filt[,i])
} 


```

#### Create Summarized Experiment Object IIa

```{r sampleAnno2a}

sample_anno2a <- data.frame(
  label = colnames(PrtnQuant2_Filt)[2:9],
  condition = c("WT", "WT", "WT", "WT", "TN", "TN", "TN", "TN"),
  replicate = c(1,2,3,4,1,2,3,4)
)

```


```{r create-se-bert-tn-a}
data_se2a <- make_se(proteins_unique = PrtnQuant2_Filt, columns = c(2:9), expdesign = sample_anno2a)


```


### Data quality checks 2a

```{r qualityChecks_bert_vs_tn_a }

plot_frequency(data_se2a)
plot_numbers(data_se2a)

```


### Normalization 2a

```{r normalize_bert_vs_tn_a}

data_se_norm2a <- normalize_vsn(data_se2a)
meanSdPlot(data_se_norm2a)

plot_normalization(data_se2a, data_se_norm2a)

```


### Filtering
```{r filter_bert_vs_tn}

data_se_norm_filt2a <- filter_missval(data_se_norm2a, thr = 0.5)
meanSdPlot(data_se_norm_filt2a)


```
### Impute missing values

```{r plot_filter_bert_vs_tn_a}


### 
plot_missval(data_se_norm2a)
plot_detect(data_se_norm2a)

###

plot_missval(data_se_norm_filt2a)
plot_detect(data_se_norm_filt2a)


```

### Imputation II a
```{r impute_bert_vs_tn_a}

# For missing not at random ('MNAR'),MinProb (QRILC)

 data_se_norm_imp2a <- impute(data_se_norm_filt2a, fun = "MinProb", q = 0.01)



# For missing at random ('MAR') data a Knn or MLE

# data_se_norm_imp2a <- impute(data_se_norm_filt2a, fun = "knn", rowmax=0.9)

 plot_imputation(data_se_norm_imp2a)

```


## Data Preprocessing for Bert vs Crispr TN1 and Bert vs Crispr TN2

#### Create Summarized Experiment Object IIb
```{r sampleAnno2b}

sample_anno2b <- data.frame(
  label = colnames(PrtnQuant2_Filt)[2:9],
  condition = c("WT", "WT", "WT", "WT", "TN1", "TN1", "TN2", "TN2"),
  replicate = c(1,2,3,4,1,2,1,2)
)

```


### Create Summarized Experiment Object II b

```{r create-se-bert-tn-b}


data_se2b <- make_se(proteins_unique = PrtnQuant2_Filt, columns = c(2:9), expdesign = sample_anno2b)


```



### Data quality checks 2b

```{r qualityChecks_bert_vs_tn_b }

plot_frequency(data_se2b)
plot_numbers(data_se2b)

```



### Normalization 2b

```{r normalize_bert_vs_tn_b}

data_se_norm2b <- normalize_vsn(data_se2b)
meanSdPlot(data_se_norm2b)

plot_normalization(data_se2b, data_se_norm2b)

```


### Filtering 2b
```{r filter_bert_vs_tn_b}

data_se_norm_filt2b <- filter_missval(data_se_norm2b, thr = 0.5)
meanSdPlot(data_se_norm_filt2b)


```

### Impute missing values 2b

```{r plot_filter_bert_vs_tn_b}


### 
plot_missval(data_se_norm2b)
plot_detect(data_se_norm2b)

###

plot_missval(data_se_norm_filt2b)
plot_detect(data_se_norm_filt2b)


```


### Imputation II b
```{r impute_bert_vs_tn_b}

# For missing not at random ('MNAR'),MinProb (QRILC)

 data_se_norm_imp2b <- impute(data_se_norm_filt2b, fun = "MinProb", q = 0.01)

# For missing at random ('MAR') data a Knn or MLE

# data_se_norm_imp2b <- impute(data_se_norm_filt2b, fun = "knn", rowmax=0.9)

 plot_imputation(data_se_norm_imp2b)

```


# Differential Enrichment Analysis

## DEP Analysis for mo_vs_tnb

```{r diff-analysis-mo-tnb}

# data_diff <- test_diff(data_se_norm_imp, type = "control", control = "Control")
# dep <- add_rejections(data_diff, alpha = 0.05, lfc = 1)


data_diff1 <- test_diff(data_se_norm_imp1, type = "control", control = "Control")
dep1 <- add_rejections(data_diff1, alpha = 0.05, lfc = 1)


```


### Heatmap for MO_vs_TNB analysis

```{r heatmap-mo-tnb}
plot_heatmap(dep1, type = "centered", kmeans = TRUE, k = 6, col_limit = 4)
```

### PCA Plot for MO_vs_TNB analysis

```{r pca-mo-tnb}
plot_pca(dep1, x = 1, y = 2, label = TRUE, n = 500)
```


## DEP Analysis for bert_vs_tn 

```{r diff-analysis-bert-tn_a}

data_diff2a <- test_diff(data_se_norm_imp2a, type = "control", control = "WT")
dep2a <- add_rejections(data_diff2a, alpha = 0.05, lfc = 1)

```

### Heatmap for bert_vs_tn analysis

```{r heatmap-bert-tn_a}

plot_heatmap(dep2a, type = "centered", kmeans = TRUE, k = 6, col_limit = 4)
```

### PCA plot for bert_vs_tn analysis

```{r pca-bert-tn_a}

plot_pca(dep2a, x = 1, y = 2, label = TRUE, n = 500)
```

## DEP Analysis for bert_vs_tn1 and bert_vs_tn2

```{r diff-analysis-bert-tn_b}

data_diff2b <- test_diff(data_se_norm_imp2b, type = "control", control = "WT")
dep2b <- add_rejections(data_diff2b, alpha = 0.05, lfc = 1)

```

### Heatmap for bert_vs_tn analysis II

```{r heatmap-bert-tn_b}

plot_heatmap(dep2b, type = "centered", kmeans = TRUE, k = 6, col_limit = 4)
```

### PCA plot for bert_vs_tn analysis II

```{r pca-bert-tn_b}
plot_pca(dep2b, x = 1, y = 2, label = TRUE, n = 500)
```


# Results

## Comparative Analysis of DEP Results 


### Volcano plot for MO_vs_TNB analysis

```{r volcano-mo-tnb}

#pVol <- plot_volcano(dep, contrast='Mutant_vs_Control', label_size=2, adjusted=TRUE, add_names=TRUE)

pVol <- plot_volcano(dep1, contrast='Mutant_vs_Control', label_size=2, adjusted=TRUE, add_names=TRUE)

# Extract data from the plot object
volcano_data <- pVol$data

# Add a new column for DE categories based on x and significance
volcano_data$DE <- with(volcano_data, 
                        ifelse(significant & x > 0, "Up", 
                               ifelse(significant & x < 0, "Down", "NS"))
)

# Convert DE column to a factor for consistent plotting
volcano_data$DE <- factor(volcano_data$DE, levels = c("Up", "Down", "NS"))

# Create the volcano plot with new DE column
pVol <- ggplot(volcano_data, aes(x = x, y = y, color = DE)) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c(
      "Up" = "red",         # Upregulated proteins
      "Down" = "blue",      # Downregulated proteins
      "NS" = "gray"         # Non-significant proteins
    )
  ) +
  geom_text_repel(
    aes(label = ifelse(DE != "NS", name, "")), # Label significant points (Up or Down)
    max.overlaps = 50,
    size = 3
  ) +
  theme_minimal() +
  labs(
    title = "Volcano Plot of DEP for MO_vs_TNB",
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value",
    color = "Expression"
  )

# Display the volcano plot
print(pVol)

# Save the customized plot

ggsave("prtnquant-volcanoplot_V2_MO_vs_TNB.png", plot = pVol, width = 10, height = 8)

```


#### Mutant Proteins LE (Glyma.02G012600), KTi3 (Glyma.08G341500), and P34 (Glyma.08G116300) genes in MO_vs_TNB comparison

```{r singlePlots}
plot_single(dep1,proteins="Glyma.02G012600",type="centered")

plot_single(dep1,proteins="Glyma.08G341500",type="centered")

plot_single(dep1,proteins="Glyma.08G116300",type="centered")

```


### Volcano plot for bert_vs_tn lines 

```{r volcano-bert-tn}

# Volcano plot
pVol2 <- plot_volcano(dep2a,contrast='TN_vs_WT',label_size=2,adjusted = TRUE, add_names = TRUE)

# Extract data from the plot object
volcano_data2 <- pVol2$data

# Add a new column for DE categories based on x and significance
volcano_data2$DE <- with(volcano_data2, 
                        ifelse(significant & x > 0, "Up", 
                               ifelse(significant & x < 0, "Down", "NS"))
)

# Convert DE column to a factor for consistent plotting
volcano_data2$DE <- factor(volcano_data2$DE, levels = c("Up", "Down", "NS"))

# Create the volcano plot with new DE column
pVol2 <- ggplot(volcano_data2, aes(x = x, y = y, color = DE)) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c(
      "Up" = "red",         # Upregulated proteins
      "Down" = "blue",      # Downregulated proteins
      "NS" = "gray"         # Non-significant proteins
    )
  ) +
  geom_text_repel(
    aes(label = ifelse(DE != "NS", name, "")), # Label significant points (Up or Down)
    max.overlaps = 50,
    size = 3
  ) +
  theme_minimal() +
  labs(
    title = "Volcano Plot of DEP for Bert_vs_TN",
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value",
    color = "Expression"
  )

# Display the volcano plot
print(pVol2)

# Save the customized plot
ggsave("prtnquant-volcanoplot_V2_Bert_vs_TN1.png", plot = pVol2, width = 10, height = 8)

###

```

#### Mutant Proteins LE (Glyma.02G012600), KTi3 (Glyma.08G341500), and P34 (Glyma.08G116300) genes in Bert_vs_CrisprTN comparison

```{r singlePlots_bert_vs_tn}
plot_single(dep2a,proteins="Glyma.02G012600",type="centered")

plot_single(dep2a,proteins="Glyma.08G341500",type="centered")

plot_single(dep2a,proteins="Glyma.08G116300",type="centered")

```


### Volcano plot for bert_vs_tn1 and bert_vs_tn2

```{r volcano-bert-tn-2b}

# Volcano plot

pVol2b <- plot_volcano(dep2b,contrast='TN1_vs_WT',label_size=2,adjusted = TRUE, add_names = TRUE)

# Extract data from the plot object
volcano_data2b <- pVol2b$data

# Add a new column for DE categories based on x and significance
volcano_data2b$DE <- with(volcano_data2b, 
                        ifelse(significant & x > 0, "Up", 
                               ifelse(significant & x < 0, "Down", "NS"))
)

# Convert DE column to a factor for consistent plotting
volcano_data2b$DE <- factor(volcano_data2b$DE, levels = c("Up", "Down", "NS"))

# Create the volcano plot with new DE column
pVol2b <- ggplot(volcano_data2b, aes(x = x, y = y, color = DE)) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c(
      "Up" = "red",         # Upregulated proteins
      "Down" = "blue",      # Downregulated proteins
      "NS" = "gray"         # Non-significant proteins
    )
  ) +
  geom_text_repel(
    aes(label = ifelse(DE != "NS", name, "")), # Label significant points (Up or Down)
    max.overlaps = 50,
    size = 3
  ) +
  theme_minimal() +
  labs(
    title = "Volcano Plot of DEP for Bert_vs_TN1",
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value",
    color = "Expression"
  )

# Display the volcano plot
print(pVol2b)

# Save the customized plot
ggsave("prtnquant-volcanoplot_V2_Bert_vs_TN1.png", plot = pVol2b, width = 10, height = 8)

###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### ###### 
###### 'TN2_vs_WT'

pVol2c <- plot_volcano(dep2b,contrast='TN2_vs_WT',label_size=2,adjusted = TRUE, add_names = TRUE)


# Extract data from the plot object
volcano_data2c <- pVol2c$data

# Add a new column for DE categories based on x and significance
volcano_data2c$DE <- with(volcano_data2c, 
                        ifelse(significant & x > 0, "Up", 
                               ifelse(significant & x < 0, "Down", "NS"))
)

# Convert DE column to a factor for consistent plotting
volcano_data2c$DE <- factor(volcano_data2c$DE, levels = c("Up", "Down", "NS"))

# Create the volcano plot with new DE column
pVol2c <- ggplot(volcano_data2c, aes(x = x, y = y, color = DE)) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c(
      "Up" = "red",         # Upregulated proteins
      "Down" = "blue",      # Downregulated proteins
      "NS" = "gray"         # Non-significant proteins
    )
  ) +
  geom_text_repel(
    aes(label = ifelse(DE != "NS", name, "")), # Label significant points (Up or Down)
    max.overlaps = 50,
    size = 3
  ) +
  theme_minimal() +
  labs(
    title = "Volcano Plot of DEP for Bert_vs_TN2",
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value",
    color = "Expression"
  )



print(pVol2c)

# Save the customized plot
ggsave("prtnquant-volcanoplot_V2_Bert_vs_TN2.png", plot = pVol2c, width = 10, height = 8)

###

```


# Conclusion
The proteins Glyma.02G012600, Glyma.08G341500, and  Glyma.08G116300, which are the target proteins in the triple null mutant and CRISPR mutant lines, are significantly down-regulated in all three comparisons with their respective controls. Further analysis of other differentially expressed proteins will be performed as part of future studies. 


```{r End}

sessionInfo()

```